Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-17 MIpaper/par_instable.tex
Revised simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@49 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Note that, as mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For example, if we suspected that the verbal factor loadings
lacked measurement invariance, we could test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.  Alternatively, we
can consider all model parameters (including means), leading to a test
of \eqref{eq:h0}.  We consider both of these tests below.

@@

\readme{I didn't look at this section in detail because I think we
should rerun the simulations first with the modified DGP.}

% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in a context
similar to the example.
We examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
tests involving only the parameters that changed with tests of all model
parameters.  The example implied that the former tests had more
power, and the simulations provide more detail on the extent to which
power increases.  Finally, we also examine the tests' power across
various magnitudes of measurement invariance violations.

\subsection{Method}
Data were generated in a similar way to that of 
the example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 13--15 years of age
having smaller factor loadings than students 16--18 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and three magnitudes of violations (factor loadings
for the younger students being $(0, 0.25, 0.5, \dots, 5)$ standard
errors\footnote{The standard errors used here were
  obtained by fitting
  the model to a sample of size 100, where the sample consisted of all
  ``older'' subjects.} below 
those for the older students).  The 0-standard error condition was
used to study Type I error rate.

For each combination of sample size $\times$
violation magnitude $\times$ number of parameters being tested,
5,000 datasets were generated and tested.  In each dataset, half the
students were 13--15 
years of age and half were 16--18 years of age.

@@

Figure~\ref{fig:simres} displays power as a function of violation
magnitude, with panels for each combination of sample size $\times$
number of parameters being tested.  
Circles reflect the double-max test, triangles reflect the max-LM
test, and plus signs reflect the Cram{\'e}r-von Mises test.  One can
generally observe that simultaneous tests of all 19 parameters
result in decreased power, with the tests performing very similarly at
the larger sample sizes.  The tests distinguish themselves from one
another when only the three factor loadings are tested, with the
Cram\'{e}r-von Mises test having the most power, followed by max-LM,
followed by the double-max test.  This advantage decreases with
increases in sample size.  
Absolute differences in power can be found in Table~\ref{tab:simres},
which contains a subset of the results.  For example, the table shows that the
power advantage of the Cram\'{e}r-von
Mises test can be as large as .1 (at $n=100$, 3 parameters being
tested).  It also shows that the Cram\'{e}r-von Mises test generally
has true Type-I error rates, with the double-max test being slightly
conservative and the max-LM test being slightly liberal at small
sample sizes.

In summary, we found that the proposed tests have adequate power to detect
measurement invariance violations in applied contexts.  The 
Cram\'{e}r-von Mises statistic exhibited the best performance for the
data generated here, though this result may not hold for all
situations.  In the discussion, we specifically consider the extent to 
which these findings may extend to other factor-analytic situations.

\begin{figure}
\caption{Simulated power for three test statistics across three sample
  sizes, two subsets of tested parameters, and measurement invariance
  violations of 0 to 5 standard 
  errors.  Circles reflect the double-max test, triangles reflect the
  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
  Numbers above each panel reflect the sample size and number of
  parameters being tested, respectively.}
\label{fig:simres}
\includegraphics[height=5in]{simres.pdf}
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes, eight magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
\label{tab:simres}
\begin{center}
\begin{tabular}{lllllllllll}
  \hline
   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
       &    2 &  2.5  &   3 &  3.5 \\ 
  \hline
  100  &  3    &  DM           &  3.1 &    3.6 &    6.6 &   12.4 &   20.6 &   32.1  &  46.1 &   57.8 \\ 
         &         &  CvM            &  5.0 &    5.7 &   10.6 &   19.1 &   32.1 &   49.2  &  67.2 &   78.8 \\ 
         &         &  max-LM          &  6.2 &    6.2 &    9.8 &   15.0 &   24.7 &   39.7  &  57.0 &   69.2 \\ 
         &  19   &  DM           &  2.5 &    2.8 &    3.1 &    4.9 &    7.1 &   12.4  &  19.6 &   25.3 \\ 
         &         &  CvM            &  4.2 &    4.8 &    5.8 &    8.6 &   13.0 &   18.9  &  28.2 &   37.4 \\ 
         &         &  max-LM          &  6.7 &    6.9 &    7.7 &    9.6 &   13.3 &   17.3  &  23.6 &   30.6 \\ 
  200  &  3    &  DM           &  3.7 &    5.2 &   12.5 &   26.3 &   47.0 &   70.6  &  86.8 &   95.0 \\ 
         &         &  CvM            &  5.3 &    7.9 &   18.6 &   39.1 &   64.3 &   85.8  &  96.2 &   99.0 \\ 
         &         &  max-LM          &  6.0 &    8.0 &   14.5 &   30.2 &   54.7 &   79.4  &  93.2 &   98.5 \\ 
         &  19   &  DM           &  3.2 &    3.6 &    5.7 &   10.8 &   21.9 &   40.2  &  61.1 &   76.5 \\ 
         &         &  CvM            &  4.6 &    5.8 &    8.5 &   15.0 &   29.1 &   46.4  &  65.4 &   80.7 \\ 
         &         &  max-LM          &  5.6 &    6.4 &    8.3 &   13.2 &   22.3 &   38.2  &  57.5 &   74.8 \\ 
  500  &  3    &  DM           &  4.1 &    8.3 &   32.2 &   69.0 &   93.4 &   99.5  & 100.0 &  100.0 \\ 
         &         &  CvM            &  4.7 &   11.8 &   42.2 &   81.3 &   98.0 &  100.0  & 100.0 &  100.0 \\ 
         &         &  max-LM          &  4.6 &    9.8 &   33.9 &   75.5 &   97.1 &  100.0  & 100.0 &  100.0 \\ 
         &  19   &  DM           &  3.6 &    4.4 &   14.0 &   40.1 &   75.3 &   96.0  &  99.7 &  100.0 \\ 
         &         &  CvM            &  5.0 &    6.7 &   17.9 &   43.6 &   77.4 &   95.9  &  99.7 &   99.9 \\ 
         &         &  max-LM          &  4.9 &    7.0 &   14.2 &   36.0 &   71.0 &   94.9  &  99.6 &  100.0 \\ 
   \hline
\end{tabular}
\end{center}
\end{table*}

>>>>

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For example, if we suspected that the verbal factor loadings
lacked measurement invariance, we could test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.  Alternatively, we
can consider all model parameters (including means), leading to a test
of \eqref{eq:h0}.  We consider both of these tests below.

@@

% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in a context
similar to the example.
We examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
tests involving only the parameters that changed with tests of all model
parameters.  The example implied that the former tests had more
power, and the simulations provide more detail on the extent to which
the tests' power levels differ.
Finally, we also examine the tests' power across
various magnitudes of measurement invariance violations.

\subsection{Method}
Data were generated in a similar way to that of 
the example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 13--15 years of age
having smaller factor loadings than students 16--18 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and fifteen magnitudes of violations.  These violations
involved the younger students' values of $\{\lambda_{11},
\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
values by $d$ 
times the parameters' asymptotic standard errors, with $d = 0, 0.25,
0.5, \dots, 3.5$.  The 0-standard error condition was
used to study Type I error rate.

For each combination of sample size ($n$) $\times$
violation magnitude ($d$) $\times$ number of parameters being tested
($k^{\ast}$),
5,000 datasets were generated and tested.  In each dataset, half the
students were 13--15 
years of age and half were 16--18 years of age.

@@

Figure~\ref{fig:simres} displays power as a function of violation
magnitude, with panels for each combination of sample size ($n$) $\times$
number of parameters being tested ($k^{\ast}$).  
Circles reflect the double-max test, triangles reflect the max-LM
test, and plus signs reflect the Cram{\'e}r-von Mises test.  One can
generally observe that simultaneous tests of all 19 parameters
result in decreased power, with the tests performing more similarly at
the larger sample sizes.  The tests distinguish themselves from one
another when only the three factor loadings are tested, with the
Cram\'{e}r-von Mises test having the most power, followed by max-LM,
followed by the double-max test.  This advantage decreases with
increases in sample size.  
Table~\ref{tab:simres} presents the same results as
Figure~\ref{fig:simres}, but it is easier to see exact power
magnitudes in the table.
The table shows that the
power advantage of the Cram\'{e}r-von
Mises test can be as large as .1, most notably when three parameters
are being tested.
  It also shows that the Cram\'{e}r-von Mises test generally
has true Type-I error rates, with the double-max test being slightly
conservative and the max-LM test being slightly liberal.

In summary, we found that the proposed tests have adequate power to detect
measurement invariance violations in applied contexts.  The 
Cram\'{e}r-von Mises statistic exhibited the best performance for the
data generated here, though more simulations are warranted to examine
the generality of this finding.
In the discussion, we describe extensions of the tests in factor
analysis and beyond.

\begin{figure}
\caption{Simulated power for three test statistics across three sample
  sizes, two subsets of tested parameters, and measurement invariance
  violations of 0--3.5 standard 
  errors (scaled by $n$).  Circles reflect the double-max test,
  triangles reflect the 
  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
  Numbers above each panel reflect the sample size and number of
  parameters being tested, respectively.}
\label{fig:simres}
\includegraphics[height=9in]{simres.pdf}
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes, fifteen magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
\label{tab:simres}
\begin{center}
\begin{tabular}{lllllllllll}
  \hline
   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
       &    2 &  2.5  &   3 &  3.5 \\ 
  \hline
100 & 3 & DM & 3.1 & 4.1 & 8.5 & 16.1 & 27.2 & 43.2 & 57.1 & 67.4 \\ 
   &  & CvM & 4.9 & 6.6 & 13.1 & 26.4 & 45.0 & 65.9 & 80.4 & 89.2 \\ 
   &  & Max-LM & 4.9 & 6.8 & 11.4 & 20.6 & 34.8 & 54.8 & 71.4 & 82.9 \\ 
   & 19 & DM & 2.4 & 3.0 & 4.3 & 5.9 & 9.5 & 15.9 & 23.8 & 31.5 \\ 
   &  & CvM & 4.6 & 4.4 & 7.0 & 11.1 & 17.1 & 27.1 & 38.1 & 49.2 \\ 
   &  & Max-LM & 7.0 & 7.1 & 9.0 & 11.2 & 16.2 & 21.4 & 31.0 & 39.7 \\ 
  200 & 3 & DM & 3.7 & 5.0 & 9.0 & 17.8 & 32.0 & 49.3 & 67.1 & 84.0 \\ 
   &  & CvM & 4.9 & 6.5 & 13.9 & 26.2 & 48.2 & 68.1 & 85.5 & 95.1 \\ 
   &  & Max-LM & 5.5 & 6.1 & 11.1 & 20.9 & 39.1 & 58.6 & 78.5 & 92.0 \\ 
   & 19 & DM & 3.8 & 3.4 & 4.6 & 7.1 & 13.1 & 22.0 & 34.7 & 52.5 \\ 
   &  & CvM & 4.6 & 4.9 & 7.3 & 11.8 & 20.1 & 31.7 & 45.8 & 64.0 \\ 
   &  & Max-LM & 6.0 & 6.1 & 8.2 & 11.3 & 16.3 & 24.4 & 37.6 & 54.3 \\ 
  500 & 3 & DM & 3.9 & 5.6 & 9.7 & 19.5 & 34.3 & 52.3 & 71.9 & 87.0 \\ 
   &  & CvM & 4.8 & 7.3 & 12.9 & 26.9 & 47.3 & 69.1 & 86.7 & 94.9 \\ 
   &  & Max-LM & 5.5 & 6.7 & 11.4 & 21.4 & 38.5 & 60.1 & 81.4 & 93.2 \\ 
   & 19 & DM & 3.7 & 4.4 & 5.3 & 8.7 & 14.6 & 25.6 & 41.6 & 60.9 \\ 
   &  & CvM & 4.8 & 5.3 & 7.1 & 11.9 & 20.6 & 32.3 & 50.3 & 68.8 \\ 
   &  & Max-LM & 5.6 & 6.2 & 7.5 & 10.4 & 15.9 & 25.8 & 41.6 & 60.8 \\ 
   \hline
\end{tabular}
\end{center}
\end{table*}
